---
title: "Movielens RMD Report"
author: "Knut Ulsrud"
date: "2/28/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load data

```{r load, echo=FALSE, include = FALSE}

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

file<- "input_data.rda"
load(file)


```

#Overview


## Goal of the project and key steps that were performed
This project's goal is to train a movie recommendation system based on Movielens data. This report is produced as a submission to the Data Science: Capstone course provided by HarvardX on the EdX platform. 

## Dataset Description
The Movielens dataset that was provided through this course has 9,000,055 rows and six columns. Columns contain the following information: 
1. userId, a unique numeric value assigned to each user u that has provided a rating
2. movieId, a unique value assigned to each movie m, in numeric format
3. rating, the rating provided to a specific movie m by a user u, in numeric format
4. timestamp, the time at which the rating was provided, in numeric format
5. . title, the name of each movie m, in character format
6. genres, the genre combinations of each movie, in character format

```{r description, echo=FALSE}

str(edx) 

```

#Mmethods and Analysis 

Process and Techniques used

##Data Cleaning
A high-level check revealed that there were no missing values in the movielens dataset provided for this assignment. 

```{r cleaning}

colSums((is.na(edx))) #There are no missing values in the dataset

#Transformed datestamp to date format and then to release year and month, and weekday. Transformed rating to integer to facilitate correlation matrix
ml_df<- edx %>% mutate(release_year = year(as.Date.POSIXct(timestamp)), 
                       release_month = month(as.Date.POSIXct(timestamp)), 
                       release_day = wday(as.Date.POSIXct(timestamp)),
                       rating = as.integer(rating))

ml_df<- ml_df %>% select(-c("movieId", timestamp)) #MovieId can be taken out because it overlaps with title. 


```

##Data Exploration and Visualization

```{r exploration_visualization}

cor(ml_df$release_month, edx$rating) #This confirms the assumption that the time of year a movie was released has little impact on rating.

correlations<- data.frame(cor(ml_df[,lapply(ml_df, class)=="integer"])) #Correlates numeric columns
correlations

ml_df %>% #indicates relationship between year and rating
  group_by(release_year) %>%
  summarize(mean_rating = mean(rating)) %>% 
  ggplot(aes(release_year, mean_rating))+
  geom_bar(stat="identity")+
  geom_smooth(method = "auto")

ml_df %>% #Reveals much fewer ratings given to movies in 1995
  group_by(release_year) %>%
  summarize(num_ratings = n()) %>% 
  ggplot(aes(release_year, num_ratings))+
  geom_bar(stat = "identity")+
  scale_y_continuous(trans = "log10")

sum(ml_df$release_year==1995)
ml_df[ml_df$release_year==1995,] #Only 2 movies released in 1995 in dataset
#Timestamp is not release year, but rating year - fix throughout

```

##Insights gained


##Modeling approach


#Results

```{r results}


```

##Appendix - null findings
```{r null}

ml_df %>% #No meaningful relationship between release month and avg. rating
  group_by(release_month) %>%
  summarize(mean_rating = mean(rating)) %>% 
  ggplot(aes(release_month, mean_rating))+
  geom_bar(stat="identity")

ml_df %>% #No meaningful relationship between release month and avg. rating
  group_by(release_day) %>%
  summarize(mean_rating = mean(rating)) %>% 
  ggplot(aes(release_day, mean_rating))+
  geom_bar(stat="identity")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
